Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-22 drafts/review-responses-april-2016.tex

% \documentclass{article}
% \usepackage{fullpage}
% \input{review-response-commands}
% 
% \begin{document}

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

@@

\begin{quote}
This is a nice analysis of how neutral genome introgresses past a cline maintained by selection against heterozygotes.  There has been a good deal of work on related problems, but mostly assuming equilibrium, and mostly on the forwards evolution of allele frequencies, rather than tracing back the ancestry of blocks of genome.  The theory presented here is therefore new, and directly relevant to interpreting genomes sampled from hybrid zones.
\end{quote}

\begin{point}{Fig 1 caption}
``alelle''
\end{point}

\reply
Fixed.

\begin{point}{\revref}
``$s$ and $\sigma$ are small'' - but relative to what?  This is not at all obvious.
\end{point}

\reply
We just meant that in the rescaling argument for a discrete model this is required formally,
without giving it enough thought.
We've added more thoughtful discussion.

\begin{point}{\revref}
Surely if time is rescaled as well as distance and genome length, everything should look the same?  
All the results, and to a good approximation the simulations, are in the continuous regime where we can rescale.
\end{point}

\reply
Good point; we in fact outlined this rescaling at \llname{ll:rescaling}.
It still seems fair to say there is ``more information'' here, because we can pick up the difference in time scaling
using haplotypes and linkage;
we've edited this paragraph to make the point better.

\begin{point}{}
A comment on two dimensions would be helpful.  
Ancestral lineages simply diffuse transversely at a rate $\sigma$, independent of their motion parallel to the cline.
\end{point}

\reply
We've added a short note on this (with more references) \revref.

\begin{point}{\revref}
This is a discrete deme model, so dispersal can't follow a continuous Gaussian. 
I guess what is happening is that $\sigma$ is chosen larger relative to deme spacing.  
However, there is some approximation here, and the approach will fail if $\sigma$ is small relatve to deme spacing.
\end{point}

\reply
There might be some confusion about our use of the floor function, which we now explicitly explain.
It's a good point more generally that if $\sigma$ is smaller than the deme spacing this scheme won't get dispersal with $\sigma$ close to the $\sigma$ that we plug into the Gaussian;
it's for this reason that we compute numerically what $\sigma$ actually is when comparing simulation to theory.

\begin{point}{}
Missing semicolon?
\end{point}

@@

\reply
Fixed.

\begin{point}{}
``are have''
\end{point}

\reply
Fixed.

\begin{point}{\revref}
``reality'' means ``simulation'' here??
\end{point}

\reply
Well, we hope to mean ``reality'', but perhaps ``verify that theory matches reality'' is too strong; we've adjusted the wording.
\alisa{I think we do mean ``reality" here, but alternate opinions welcome}
\plr{I changed to ``realistic, individual-based models''.}

@@

\begin{point}{}
 Some appendices might be transferred to supplementary material -- the derivations are fairly standard.
\end{point}

\reply
\alisa{Our decision to keep the derivations in appendices was with consideration for the readership of Molecular Ecology}
\plr{I'm fine with moving some; not super clear on the difference anyhow...}

### 2016-04-22 drafts/hybrid_zone.tex
went through AE comments

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, Barton1986}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
(iii) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
and (iv) population density is large,
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) .
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian.
\plr{change $\dot p$ to $\frac{d}{dt}p$, and $\Delta$ to the second derivative?}
This says that the local rate of change of the frequency of $A$ ($\dot p$)
is equal to the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$)
and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
due to local proportions of heterozygotes,
which is $s(2p-1)$ (up to an error of size $s^2$).
This can be derived as in \citet{Nagylaki1975};
for discussion of diploidy see \citet{christiansen1995genotypic}.
As we show below the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation \citep{Slatkin1973},
implying that the cline establishes over a timescale of the order $1/s$. 
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

@@

Note that this description gives more information than the steady-state cline $p(x)$,
which depends only on $\sigma/\sqrt{s}$.
Here we see that lineages with $\sigma=10$ and $s=.16$ move quite differently
to lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form.

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
\alisa{Use our words}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian \alisa{describing the rate of change}. 
The three terms in these equations come from Brownian movement of a linages,
the tendency of a lineage to inherit from regions where its type is more common,
and recombination between selected backgrounds, respectively.
Equivalently,
the first term ($\Delta q$) reflects the smoothing action of local dispersal,
the second is the net flux induced by reduced fitness in the hybrid zone,
and the third is the contribution of recombination.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$G^A = \sigma^2 \grad \log(p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$
and
$G^B = \sigma^2 \grad \log(1-p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$,
so that equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \dot q_A &= G^A q_A + r (1-p) (q_B-q_A) \\
    \dot q_B &= G^B q_B + r p (q_A-q_B) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of lineages of selected alleles of ancestry $A$ and $B$, respectively;
informally, they encode the stochastic motion of lineages at the selected loci.

@@

The resulting equation is similar to \eqref{eqn:q_pde} with a term added for branching,
which is written (omitting the $(x,t)$ for conciseness):
\begin{align}
    \begin{aligned} \label{eqn:g_pde}
        \dot g_A(a,b) 
        &= G^A g_A(a,b) + r(a,b) (1-p) (g_A(a,b)-g_B(a,b))
            \\ {} & \qquad 
            + (1-p) \left( 
                \int_a^{\min(0,b)} g_B(a,\theta) g_A(\theta,b) d\theta
                + \int_{\max(0,a)}^b g_A(a,\theta) g_B(\theta,b) d\theta 
            \right)
            \\ {} & \qquad 
            + p \int_a^b {
                g_A(a,\theta) g_A(\theta,b) 
            } d\theta
            - (b-a) g_A(a,b)  ,
    \end{aligned} 
\end{align}

@@

Each deme has exactly $N_d$ diploid hermaphroditic individuals at the start of each generation.
Then, every individual can migrate to a new deme  offset by $\lfloor Z + 1/2 \rfloor$ from their original position,
where $Z$ is a Gaussian random variable with mean zero and variance $\sigma^2$.
(The mean displacement is zero, and when comparing to theory we compute $\sigma$ as the standard deviation of this distribution.)
Migrants past either end of the range remain at the terminal demes.
Then, 
fitnesses are computed (heterozygotes at the selected locus have fitness $1-s$; homozygotes have fitness 1),
and in each deme $N_d$ pairs of parents are chosen, with replacement,
with probability proportional to their fitness and selfing allowed.
Since migration is not conservative, demes may have no available parents;
in this case, parents are chosen from other demes with probability proportional to fitness
multiplied by $\exp(-x^2/\sigma^2)$, where $x$ is the distance to the other deme.
The next generation is formed by carrying out meiosis in each parent
and combining the resulting gametes such that each pair of parents leaves one descendant.
Meiosis results in alternating blocks of the gamete's chromosome
being inherited from the two parental chromosomes,
with the blocks separated by a Poisson(1) number of uniformly chosen recombination points along the chromosome,
and the order of the parental chromosomes chosen randomly.
The simulation software works by recording, for each chromosome, 
a list of ancestry breakpoints, and the index of the ancestor from which 
the chromosome inherited that genomic region.
We then assigned ancestry at individual loci by looking up which side of the zone the ancestor lived on.
% removed "ancestor's genotype" because it's not really genotype

@@

The diffusion model for lineages predicts that quantities of interest solve sets of coupled partial differential equations (PDE),
which we have written down.
As there are no known analytical solutions to these PDE,
we have constructed numerical solutions (and provide the source code for doing this).
A main role of these solutions in our work has been to verify that theory based on the diffusion model of lineages
matches reality.
These solutions easily and quickly provide predictions of joint frequencies at small numbers of loci:
about 1 second to compute predicted clines, as opposed to hours for the full simulation.
However, due to the high dimensionality of the haplotype problem (spatial position $\times$ time $\times$ endpoints of the haplotype),
numerical solutions for mean haplotype lengths along the genome can be as computationally intensive as simulations
(although are substantially less noisy).
More work could be done to develop more efficient methods of solution, 
but it may be better to perform more biologically realistic forwards-time simulations that includes coalescence and drift to better characterize the block length distribution.
If however the correlation in ancestry, rather than the block-length distribution, is of interest the PDE approach may be preferable because it is easily modified to provide predictions for spatial and temporally inhomogeneous systems -- for instance, across maps of real landscapes.

@@

%% response to reviews
% to remove these, change 'ifreviewresponses' up top
\includereviews

>>>>

\plr{Need to cite Durrett and Cox's stuff here, especially \citet{durrett2007width}.}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, Barton1986}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\Delta p$ is the Laplacian.
\plr{change $\Delta$ to the second derivative?}
The equation is approximate because it 
omits terms of order $s^2$
(so deviations from the prediction occur over $1/s^2$ generations),
and relies on the Central Limit Theorem to approximate generation-to-generation dispersal by the Gaussian
(so if dispersal is non-Gaussian, will fit best over longer time scales, say, tens of generations).  \revpoint{AE}{2}
This says that the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$)
is equal to the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$)
and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
due to local proportions of heterozygotes,
which is $s(2p-1)$ (up to an error of size $s^2$).
This can be derived as in \citet{Nagylaki1975};
for discussion of diploidy see \citet{christiansen1995genotypic}.
As we show below the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation \citep{Slatkin1973}, \linelabel{ll:rescaling}
implying that the cline establishes over a timescale of the order $1/s$. 
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

For simplicity we mostly work with one-dimensional clines,
but the mathematics applies as well to two-dimensional systems with some modifications:
for instance, if the hybrid zone is a straight line, 
the description above applies to motion of a lineage transverse to the hybrid zone.  \revpoint{AE}{4}
It is for this reason that we use $\Delta$ for the second derivative and $\grad$ for the first derivative,
as these carry over without change to two dimensions;
in fact, if the landscape is heterogeneous or dispersal is non-Gaussian,
the theory still holds replacing $\Delta$ by the appropriate operator.
Note, however, that in two dimensions drift can have a much stronger effect \citep{cox1995hybrid,barton2010phylogeography,durrett2007width}.

@@

This description gives more information than the steady-state cline $p(x)$,
which depends only on $\sigma/\sqrt{s}$.
Here we see that lineages with $\sigma=10$ and $s=.16$ move much faster
than lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form.
Even though a rescaling of space and time as described above can make the models equivalent,
this difference in lineage speed can be seen through the action of recombination, which we explore next.
\revpoint{AE}{3}

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
\alisa{Use our words}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian \alisa{describing the rate of change}. 
The three terms in these equations come from Brownian movement of a linages,
the tendency of a lineage to inherit from regions where its type is more common,
and recombination between selected backgrounds, respectively.
Equivalently,
the first term ($\Delta q$) reflects the smoothing action of local dispersal,
the second is the net flux induced by reduced fitness in the hybrid zone,
and the third is the contribution of recombination.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$G^A = \sigma^2 \grad \log(p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$
and
$G^B = \sigma^2 \grad \log(1-p(x,t)) \cdot \grad + (\sigma^2/2) \Delta$,
so that equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \frac{d}{dt} q_A &= G^A q_A + r (1-p) (q_B-q_A) \\
    \frac{d}{dt} q_B &= G^B q_B + r p (q_A-q_B) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of lineages of selected alleles of ancestry $A$ and $B$, respectively;
informally, they encode the stochastic motion of lineages at the selected loci.

@@

The resulting equation is similar to \eqref{eqn:q_pde} with a term added for branching,
which is written (omitting the $(x,t)$ for conciseness):
\begin{align}
    \begin{aligned} \label{eqn:g_pde}
        \frac{d}{dt} g_A(a,b) 
        &= G^A g_A(a,b) + r(a,b) (1-p) (g_A(a,b)-g_B(a,b))
            \\ {} & \qquad 
            + (1-p) \left( 
                \int_a^{\min(0,b)} g_B(a,\theta) g_A(\theta,b) d\theta
                + \int_{\max(0,a)}^b g_A(a,\theta) g_B(\theta,b) d\theta 
            \right)
            \\ {} & \qquad 
            + p \int_a^b {
                g_A(a,\theta) g_A(\theta,b) 
            } d\theta
            - (b-a) g_A(a,b)  ,
    \end{aligned} 
\end{align}

@@

Each deme has exactly $N_d$ diploid hermaphroditic individuals at the start of each generation.
Then, every individual can migrate to a new deme offset by $\lfloor Z + 1/2 \rfloor$ from their original position,
where $Z$ is a Gaussian random variable with mean zero and variance $\sigma^2$.
(Here $\lfloor z \rfloor$ is the largest integer less than or equal to $z$,
so this equivalently takes a centered Gaussian displacement and assigns it to the nearest deme. \revpoint{AE}{5}
The mean displacement is zero, and when comparing to theory we compute $\sigma$ as the standard deviation of this distribution.)
Migrants past either end of the range remain at the terminal demes.
Then, 
fitnesses are computed (heterozygotes at the selected locus have fitness $1-s$; homozygotes have fitness 1),
and in each deme $N_d$ pairs of parents are chosen, with replacement,
with probability proportional to their fitness and selfing allowed.
Since migration is not conservative, demes may have no available parents;
in this case, parents are chosen from other demes with probability proportional to fitness
multiplied by $\exp(-x^2/\sigma^2)$, where $x$ is the distance to the other deme.
The next generation is formed by carrying out meiosis in each parent
and combining the resulting gametes such that each pair of parents leaves one descendant.
Meiosis results in alternating blocks of the gamete's chromosome
being inherited from the two parental chromosomes,
with the blocks separated by a Poisson(1) number of uniformly chosen recombination points along the chromosome,
and the order of the parental chromosomes chosen randomly.
The simulation software works by recording, for each chromosome, 
a list of ancestry breakpoints, and the index of the ancestor from which 
the chromosome inherited that genomic region.
We then assigned ancestry at individual loci by looking up which side of the zone the ancestor lived on.
% removed "ancestor's genotype" because it's not really genotype

@@

The diffusion model for lineages predicts that quantities of interest solve sets of coupled partial differential equations (PDE),
which we have written down.
As there are no known analytical solutions to these PDE,
we have constructed numerical solutions (and provide the source code for doing this).
A main role of these solutions in our work has been to verify that theory based on the diffusion model of lineages
matches realistic, individual-based models.  \revpoint{AE}{8}
These solutions easily and quickly provide predictions of joint frequencies at small numbers of loci:
about 1 second to compute predicted clines, as opposed to hours for the full simulation.
However, due to the high dimensionality of the haplotype problem (spatial position $\times$ time $\times$ endpoints of the haplotype),
numerical solutions for mean haplotype lengths along the genome can be as computationally intensive as simulations
(although are substantially less noisy).
More work could be done to develop more efficient methods of solution, 
but it may be better to perform more biologically realistic forwards-time simulations that includes coalescence and drift to better characterize the block length distribution.
If however the correlation in ancestry, rather than the block-length distribution, is of interest the PDE approach may be preferable because it is easily modified to provide predictions for spatial and temporally inhomogeneous systems -- for instance, across maps of real landscapes.

@@

%% response to reviews
% to remove these, change 'ifreviewresponses' up top
\singlespacing
\includereviews
